---
output:
  pdf_document: default
  html_document: default
---
#install packages
```{r}
install.packages('readr')
install.packages('dplyr')
install.packages('ggplot2')
install.packages('scales')
install.packages('lubridate')
install.packages('zoo')
install.packages('ggrepel')
```

#load libraries
```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(scales)
library(lubridate)
library(zoo)
library(ggrepel)
```



#load data
```{r}
sourcing_data <- read_csv("Procurement_Cleaned_Data.csv")
```
#view data
```{r}
head(sourcing_data)
```

#structure of the data
```{r}
str(sourcing_data)
```
#change the dates to date datatype
```{r}
sourcing_data$Delivery_Date <- as.Date(sourcing_data$Delivery_Date, format = "%m/%d/%Y")
sourcing_data$Order_Date <- as.Date(sourcing_data$Order_Date, format = "%m/%d/%y")
```

#verify the datatype change
```{r}
str(sourcing_data$Delivery_Date)
summary(sourcing_data$Delivery_Date)

str(sourcing_data$Order_Date)
summary(sourcing_data$Order_Date)
```

#column names
```{r}
names(sourcing_data)
```

#summarize $ amount purchased by supplier by year

```{r}
summary_data <- sourcing_data %>%
  group_by(Supplier, Delivery_Year) %>%
  summarise(Total_Purchase_Amount = sum(Total_Purchase_Amount, na.rm = TRUE)) %>%
  ungroup()
```
#identify desc order of suppliers based on 2023 spend
```{r}
top_2023_suppliers <- summary_data %>%
  filter(Delivery_Year == 2023) %>%
  arrange(desc(Total_Purchase_Amount)) %>%
  pull(Supplier)
```


#reorder supplier factor levels
```{r}
summary_data <- summary_data %>%
  mutate(Supplier = factor(Supplier, levels = top_2023_suppliers))
```


#plot the graph
```{r}
ggplot(summary_data, aes(x = Supplier, y = Total_Purchase_Amount, fill = as.factor(Delivery_Year))) + 
  geom_bar(stat = "identity", position = "dodge") + 
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
  labs(title = "Total Purchase Amount by Supplier and Year", 
       x = "Supplier",
       y = "Total Purchase Amount", 
       fill = "Delivery Year") + 
  theme_minimal()
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#summarize $ amount purchased by category by year

```{r}
summary_category_data <- sourcing_data %>%
  group_by(Item_Category, Delivery_Year) %>%
  summarise(Total_Purchase_Amount = sum(Total_Purchase_Amount, na.rm = TRUE)) %>%
  ungroup()
```
#identify desc order of category based on 2023 spend
```{r}
top_2023_categories <- summary_category_data %>%
  filter(Delivery_Year == 2023) %>%
  arrange(desc(Total_Purchase_Amount)) %>%
  pull(Item_Category)
```
#reorder supplier factor levels
```{r}
summary_category_data <- summary_category_data %>%
  mutate(Item_Category = factor(Item_Category, levels = top_2023_categories))
```




#plot the graph
```{r}
ggplot(summary_category_data, aes(x = Item_Category, y = Total_Purchase_Amount, fill = as.factor(Delivery_Year))) + 
  geom_bar(stat = "identity", position = "dodge") + 
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
  labs(title = "Total Purchase Amount by Category and Year", 
       x = "Category",
       y = "Total Purchase Amount", 
       fill = "Delivery Year") + 
  theme_minimal()
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#summarize $ amount saved by supplier by year
```{r}
summary_savings_data <- sourcing_data %>%
  group_by(Supplier, Delivery_Year) %>%
  summarise(Negotiated_Savings = sum(Negotiated_Savings, na.rm = TRUE)) %>%
  ungroup()
```

#identify desc order of savings by supplier based on 2023 spend
```{r}
top_2023_savings <- summary_savings_data %>%
  filter(Delivery_Year == 2023) %>%
  arrange(desc(Negotiated_Savings)) %>%
  pull(Supplier)
```

#reorder supplier factor levels
```{r}
summary_savings_data <- summary_savings_data %>%
  mutate(Supplier = factor(Supplier, levels = top_2023_savings))
```

#plot the graph
```{r}
ggplot(summary_savings_data, aes(x = Supplier, y = Negotiated_Savings, fill = as.factor(Delivery_Year))) + 
  geom_bar(stat = "identity", position = "dodge") + 
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
  labs(title = "Total Negotiated Savings Amount by Supplier and Year", 
       x = "Supplier",
       y = "Negotiated Savings", 
       fill = "Delivery Year") + 
  theme_minimal()
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#monthly summary comparison
```{r}
monthly_summary <- sourcing_data %>%
  group_by(Delivery_Year, Delivery_Month) %>%
  summarise(Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE), Orders = n()) %>%
  arrange(Delivery_Year, Delivery_Month)
```





#bar chart monthly spend
```{r}
# Prepare data
monthly_summary <- sourcing_data %>%
  mutate(Delivery_MonthDate = floor_date(Delivery_Date, unit = "month")) %>%
  group_by(Delivery_MonthDate) %>%
  summarise(
    Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE),
    Orders = n()
  ) %>%
  arrange(Delivery_MonthDate) %>%
  ungroup() %>%
  # Create labels for x-axis:
  mutate(
    MonthLabel = ifelse(
      month(Delivery_MonthDate) == 1,
      paste0("Jan ", year(Delivery_MonthDate)),
      format(Delivery_MonthDate, "%b")  # This gives "Feb", "Mar", etc.
    )
  )

# Plot with explicit breaks and labels:
ggplot(monthly_summary, aes(x = Delivery_MonthDate, y = Total_Spend)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M", prefix = "$")) +
  scale_x_date(
    breaks = monthly_summary$Delivery_MonthDate,  # exact breaks
    labels = monthly_summary$MonthLabel,          # exact labels
    expand = c(0, 0)
  ) +
  labs(title = "Monthly Spend Totals",
       x = NULL,
       y = "Total Spend") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#line graph for monthly data to compare 2022 to 2023
```{r}
# Prepare data: get year, month number, and total spend per month-year
monthly_summary <- sourcing_data %>%
  mutate(
    Delivery_Year = year(Delivery_Date),
    Delivery_Month = month(Delivery_Date)
  ) %>%
  group_by(Delivery_Year, Delivery_Month) %>%
  summarise(
    Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE),
    Orders = n()
  ) %>%
  ungroup()

# Create factor for month to ensure Jan-Dec order
monthly_summary <- monthly_summary %>%
  mutate(
    Month_Factor = factor(
      month.abb[Delivery_Month], levels = month.abb
    )
  )

# Plot line chart
ggplot(monthly_summary, aes(x = Month_Factor, y = Total_Spend, group = factor(Delivery_Year), color = factor(Delivery_Year))) +
  geom_line(linewidth = 1.2) +      # changed size to linewidth here
  geom_point(size = 2) +             # size is still used for points
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M", prefix = "$")) +
  labs(
    title = "Monthly Spend by Year",
    x = "Month",
    y = "Total Spend",
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top"
  )
```
#rolling 3 month average chart
```{r}
# Step 1: Create a date column for grouping by month
monthly_summary <- sourcing_data %>%
  mutate(Delivery_MonthDate = floor_date(as.Date(Delivery_Date), "month")) %>%
  group_by(Delivery_MonthDate) %>%
  summarise(
    Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE),
    Orders = n(),
    .groups = "drop"
  ) %>%
  arrange(Delivery_MonthDate)

# Step 2: Rolling average over full timeline
monthly_summary <- monthly_summary %>%
  mutate(Rolling_Spend = rollmean(Total_Spend, k = 3, fill = NA, align = "right")) %>%
  mutate(
    Delivery_Year = year(Delivery_MonthDate),
    Delivery_Month = month(Delivery_MonthDate),
    Month_Factor = factor(month.abb[Delivery_Month], levels = month.abb)
  )

# Step 3: Plot rolling average
ggplot(monthly_summary, aes(x = Month_Factor, y = Rolling_Spend, group = factor(Delivery_Year), color = factor(Delivery_Year))) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M", prefix = "$")) +
  labs(
    title = "Rolling 3-Month Average Spend by Year",
    x = "Month",
    y = "Rolling 3-Month Avg Spend",
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top"
  )
```
#overlaying the 3 month trend and actual
```{r}
# Step 1: Summarize monthly spend and compute rolling average
monthly_summary <- sourcing_data %>%
  mutate(Delivery_MonthDate = floor_date(as.Date(Delivery_Date), "month")) %>%
  group_by(Delivery_MonthDate) %>%
  summarise(
    Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE),
    Orders = n(),
    .groups = "drop"
  ) %>%
  arrange(Delivery_MonthDate) %>%
  mutate(
    Rolling_Spend = rollmean(Total_Spend, k = 3, fill = NA, align = "right"),
    Delivery_Year = year(Delivery_MonthDate),
    Delivery_Month = month(Delivery_MonthDate),
    Month_Factor = factor(month.abb[Delivery_Month], levels = month.abb)
  )

# Step 2: Plot both actual spend and rolling average
ggplot(monthly_summary, aes(x = Month_Factor)) +
  # Bars for monthly spend
  geom_col(aes(y = Total_Spend, fill = factor(Delivery_Year)), position = position_dodge(width = 0.8), width = 0.7, alpha = 0.6) +
  # Line for rolling 3-month average
  geom_line(aes(y = Rolling_Spend, group = factor(Delivery_Year), color = factor(Delivery_Year)), linewidth = 1.2) +
  geom_point(aes(y = Rolling_Spend, color = factor(Delivery_Year)), size = 2) +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M", prefix = "$")) +
  labs(
    title = "Monthly Spend with Rolling 3-Month Average",
    x = "Month",
    y = "Spend",
    fill = "Year",
    color = "Rolling Avg Year"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top"
  )
```



#percent savings by supplier, not dollar value savings next
#summarize both savings and spend by supplier and year
```{r}
summary_savings_data <- sourcing_data %>%
  group_by(Supplier, Delivery_Year) %>%
  summarise(
    Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE),
    Negotiated_Savings = sum(Negotiated_Savings, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Savings_Percent = Negotiated_Savings / Total_Spend)
```

#order suppliers by 2023 savings percentage
```{r}
top_2023_savings <- summary_savings_data %>%
  filter(Delivery_Year == 2023) %>%
  arrange(desc(Savings_Percent)) %>%
  pull(Supplier)

summary_savings_data <- summary_savings_data %>%
  mutate(Supplier = factor(Supplier, levels = top_2023_savings))
```

#plot percentage savings by supplier
```{r}
library(ggplot2)
library(scales)

ggplot(summary_savings_data, aes(x = Supplier, y = Savings_Percent, fill = as.factor(Delivery_Year))) + 
  geom_bar(stat = "identity", position = "dodge") + 
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Negotiated Savings by Supplier and Year", 
    x = "Supplier",
    y = "Percent Saved", 
    fill = "Delivery Year"
  ) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#defect 2nd attempt - defect value based on compliant orders
```{r}
# Summarize compliance vs. defects
compliance_summary <- sourcing_data %>%
  group_by(Delivery_Year, Compliance) %>%
  summarise(
    Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE),
    Total_Defects = sum(Defect_Value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Defect_Rate = Total_Defects / Total_Spend)

# Plot
ggplot(compliance_summary, aes(x = factor(Delivery_Year), y = Defect_Rate, fill = Compliance)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = percent(Defect_Rate, accuracy = 0.1)),
            position = position_dodge(width = 0.9),
            vjust = -0.3, size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1), expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("Yes" = "#00C400", "No" = "#D00000")) +
  labs(
    title = "Defect Rate by Year and Compliance Status",
    x = "Year",
    y = "Defect Rate (% of Spend)",
    fill = "Compliance"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(face = "bold"),
    legend.position = "top"
  )
```
#Defect Rate Per Supplier
```{r}
defect_summary <- sourcing_data %>%
  group_by(Supplier, Delivery_Year) %>%
  summarise(Total_Defects = sum(Defect_Value, na.rm = TRUE)) %>%
  ungroup()

top_2023_defects <- defect_summary %>%
  filter(Delivery_Year == 2023) %>%
  arrange(desc(Total_Defects)) %>%
  pull(Supplier)

defect_summary <- defect_summary %>%
  mutate(Supplier = factor(Supplier, levels = top_2023_defects))

ggplot(defect_summary, aes(x = Supplier, y = Total_Defects, fill = as.factor(Delivery_Year))) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
  labs(
    title = "Total Defect Value by Supplier and Year",
    x = "Supplier",
    y = "Defect Value ($)",
    fill = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```
#defect rate by supplier, with percent of spend
```{r}
defect_summary <- sourcing_data %>%
  group_by(Supplier, Delivery_Year) %>%
  summarise(
    Total_Defects = sum(Defect_Value, na.rm = TRUE),
    Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE)
  ) %>%
  mutate(Defect_Rate = Total_Defects / Total_Spend * 100) %>%
  ungroup()

top_2023_defects <- defect_summary %>%
  filter(Delivery_Year == 2023) %>%
  arrange(desc(Defect_Rate)) %>%
  pull(Supplier)

defect_summary <- defect_summary %>%
  mutate(Supplier = factor(Supplier, levels = top_2023_defects))

ggplot(defect_summary, aes(x = Supplier, y = Defect_Rate, fill = as.factor(Delivery_Year))) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Defect Rate by Supplier and Year",
    x = "Supplier",
    y = "Defect Rate (%)",
    fill = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#defect rate with percentage of spend and totals
```{r}
scale_factor <- max(defect_summary$Total_Defects) / max(defect_summary$Defect_Rate)

ggplot(defect_summary, aes(x = Supplier, y = Total_Defects)) +
  geom_col(aes(fill = as.factor(Delivery_Year)), position = position_dodge(width = 0.8), width = 0.6) +
  
  geom_line(
    aes(y = Defect_Rate * scale_factor, group = as.factor(Delivery_Year), color = as.factor(Delivery_Year)),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  
  scale_y_continuous(
    name = "Total Defect Value ($)",
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    sec.axis = sec_axis(
      ~ . / scale_factor,
      name = "Defect Rate (%)",
      labels = scales::percent_format(scale = 1)
    )
  ) +
  
  labs(
    title = "Defect Value and Defect Rate by Supplier and Year",
    x = "Supplier",
    fill = "Year (Bars)",
    color = "Year (Line)"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```
#defect Rate by category by year
```{r}
# Summarize by Category and Year
defect_summary <- sourcing_data %>%
  group_by(Item_Category, Delivery_Year) %>%
  summarise(
    Total_Defects = sum(Defect_Value, na.rm = TRUE),
    Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Defect_Rate = Total_Defects / Total_Spend * 100)

# Get order based on most recent year's defect rate (e.g., 2023)
category_order <- defect_summary %>%
  filter(Delivery_Year == 2023) %>%
  arrange(desc(Defect_Rate)) %>%
  pull(Item_Category)

# Apply factor order to control bar sort
defect_summary <- defect_summary %>%
  mutate(Item_Category = factor(Item_Category, levels = category_order))

# Compute scale factor to align percent with dollar values
scale_factor <- max(defect_summary$Total_Defects) / max(defect_summary$Defect_Rate)

# Add scaled y-position for line and labels
defect_summary <- defect_summary %>%
  mutate(
    scaled_rate_y = Defect_Rate * scale_factor,
    label_y = scaled_rate_y + max(Total_Defects) * 0.05  # small space above the line
  )

# Plot
ggplot(defect_summary, aes(x = Item_Category, y = Total_Defects)) +
  geom_col(aes(fill = as.factor(Delivery_Year)),
           position = position_dodge(width = 0.8),
           width = 0.6) +

  geom_line(
    aes(y = scaled_rate_y, group = as.factor(Delivery_Year), color = as.factor(Delivery_Year)),
    position = position_dodge(width = 0.8),
    linewidth = 1
  ) +

  # Add percentage labels
#  geom_text(
#    aes(y = label_y, label = paste0(round(Defect_Rate, 1), "%"), group = as.factor(Delivery_Year), color = #as.factor(Delivery_Year)),
#    position = position_dodge(width = 0.8),
#    size = 3,
#    vjust = 0
#  ) +

  scale_y_continuous(
    name = "Total Defect Value ($)",
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    sec.axis = sec_axis(
      ~ . / scale_factor,
      name = "Defect Rate (%)",
      labels = scales::percent_format(scale = 1)
    )
  ) +

  labs(
    title = "Defect Value and Defect Rate by Item Category and Year",
    x = "Item Category",
    fill = "Year (Bars)",
    color = "Year (Line)"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```
#compliance status by supplier by year
```{r}
# Summarize total spend by Supplier, Year, and Compliance
compliance_summary <- sourcing_data %>%
  group_by(Supplier, Delivery_Year, Compliance) %>%
  summarise(Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE), .groups = "drop")

# Calculate percent of spend that is compliant per supplier and year
compliance_summary <- compliance_summary %>%
  group_by(Supplier, Delivery_Year) %>%
  mutate(Spend_Share = Total_Spend / sum(Total_Spend)) %>%
  ungroup()

# Order suppliers by non-compliant spend share in most recent year (e.g., 2023)
supplier_order <- compliance_summary %>%
  filter(Delivery_Year == 2023, Compliance == "No") %>%
  arrange(desc(Spend_Share)) %>%
  pull(Supplier)

# Apply factor order
compliance_summary <- compliance_summary %>%
  mutate(Supplier = factor(Supplier, levels = unique(supplier_order)))

# Plot
ggplot(compliance_summary, aes(x = Supplier, y = Spend_Share, fill = Compliance)) +
  geom_col(position = "stack", width = 0.7) +
  facet_wrap(~Delivery_Year, nrow = 1) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    name = "Share of Total Spend"
  ) +
  scale_fill_manual(
    values = c("No" = "#d73027", "Yes" = "#1a9850"),
    name = "Compliance"
  ) +
  labs(
    title = "Compliance Status by Supplier and Year",
    x = "Supplier"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```
#compliance status by supplier by year
```{r}
# Summarize total spend by Item_Category, Year, and Compliance
compliance_summary <- sourcing_data %>%
  group_by(Item_Category, Delivery_Year, Compliance) %>%
  summarise(Total_Spend = sum(Total_Purchase_Amount, na.rm = TRUE), .groups = "drop")

# Calculate percent of spend that is compliant per supplier and year
compliance_summary <- compliance_summary %>%
  group_by(Item_Category, Delivery_Year) %>%
  mutate(Spend_Share = Total_Spend / sum(Total_Spend)) %>%
  ungroup()

# Order suppliers by non-compliant spend share in most recent year (e.g., 2023)
supplier_order <- compliance_summary %>%
  filter(Delivery_Year == 2023, Compliance == "No") %>%
  arrange(desc(Spend_Share)) %>%
  pull(Item_Category)

# Apply factor order
compliance_summary <- compliance_summary %>%
  mutate(Item_Category = factor(Item_Category, levels = unique(supplier_order)))

# Plot
ggplot(compliance_summary, aes(x = Item_Category, y = Spend_Share, fill = Compliance)) +
  geom_col(position = "stack", width = 0.7) +
  facet_wrap(~Delivery_Year, nrow = 1) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    name = "Share of Total Spend"
  ) +
  scale_fill_manual(
    values = c("No" = "#d73027", "Yes" = "#1a9850"),
    name = "Compliance"
  ) +
  labs(
    title = "Compliance Status by Category and Year",
    x = "Category"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

